### 2022-03-24 rmd/manuscript.Rmd
edit: manuscript

---
Title: "Ecosystem size and complexity are extrinsic drivers of food chain length in branching networks"
output:
  pdf_document:
    latex_engine: xelatex
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
---

@@

We generatedsimulated branching networks of connected habitat patches as a series of nodes (Figure 1). The geometric arrangement of networks was controlled with two parameters: ecosystem size, the number of patches (NpatchNP); and ecosystem complexity, branching probability (PbB). Nodes (habitat patches) were assigned to be a confluence (or upstream terminal node) with probability PbB, or an in-branch node with probability 1-Pb 1 - PB. Branches (tributaries) are a series of connected nodes between confluences (or upstream terminal nodes). The number of patches within a branch,np, q q, is a random variable drawn from a geometric distribution np,q Ge(Pb)np,q∼Ge(Pb)  (a geometric distribution is the discrete form of an exponential distribution) but conditional on q=1Nbnp,q=Np (Nb is the number of branches). The number of branches NbB Under this assumption, the number of branches is identical to the number of branching nodes or the number of "success" out of Npatch trials. This translates into the assumption that the variable Nb follows a Binomial distribution asis a random variable drawn from a Binomial distribution: NbBinomial(Npatch, Pb).NbB \~ Binomial(NP, PbB). Generated river networks preserve important geometric scaling relationships (i.e., Horton's law). This framework preserves the fractal nature of branching patterns. For each scenario described below, we generated 1000 networks across a gradient of ecosystem size and complexity by drawing NP and Pba random variable from uniform distributions as: NpatchUnif(5, 150) NP \~ Unif(5, 150) and PbUnif(0.01, 0.99)PbB \~ Unif(0.01, 0.99) (Np was rounded to the nearest integer before running simulations).

For each terminal branch (i.e., headwater branch), h, within the network, a disturbance value, mh, was sampled .  The distribution of mh values was sampled from a normal distribution in a logit scale. This was accomplished by transforming the proportional mean disturbance value, µm, to the logit scale and sampling a value for each upstream terminal node (headwater) as logit mh Normal(logit m, m2).mh \~ Normal( logit( µm), σ2m). Here, we varied the mean source disturbance value, m = (0.1, 0.9) µm = (0.1, 0.9), based on disturbance regime (see **Simulations**) and set the variation of source disturbance as mσ2m = 2. We added lLocal environmental noise (i.e., noise occuring at each node) towas included in the disturbance value of each node as a spatial autoregressive process: logit mdown = logit mup+. mdown and mup are the disturbance values at down- and upstream nodes, and the local environmental noise at the downstream node that is drawn from a normal distribution as Normal(0, 2). This autoregressive process resembles natural disturbance regime, where flood or drought disturbance cascades downstrem with some local variation along the channel.

s moving downstream by sampling a random variable from Normal(0, σ2l). For the main simulation results presented here, σ2l = 0.01. At confluence nodes, the disturbance value takes a weighted mean based on the relative size of the two contributing branches, s (the number of upstream contributing nodespatches): logit mdown= logit m1,up+(1-) logit m2,upmdown = ωm1,up + (1 - ω)m2,up, where mdownmdown is the disturbance magnitude of the focal patch (the confluence), m1,upm1, up and m2,upm2, up are the disturbance magnitude of upstream adjacentcontributing nodestributaries, and ω the relative size of one tributary over another defined as =s1/(s1+s2). = s1/(s1 + s2). Disturbance value distributions had medians close to µm with long tails (**Fig S1**). with long right or left tails (**Fig S1**), respectively, i.e. For example, when m=0.1µm = 0.1 (e.g., a 10% reduction in species densities), most sites experienced low-magnitude disturbances and a few sites experienced high-magnitude disturbances. Likewise, whenm=0.9,  µm = 0.9, most sites experienced high-magnitude disturbances with a few patches receiving low-magnitude disturbances. The low-magnitude patches can be viewed as refugia and a source to supply individuals to colonize impacted sites in subsequent timesteps. Example disturbance value distributions for simulation scenarios reported here are available in the SI.

Carrying capacity in patch x , Kx ,was assumed to increase as a power function of ecosystem area (i.e., watershed area)58:

Kx= Kbase+C\*Npatch, up

Where Kbase=100, C is a constant, Npatch, up number of patches upstream, and a random variable sampled from Unif(1.1, 1.2). At upstream terminal nodes (headwaters) K=Kbase. The R package mcbrnet was used to simulate the networks (Citation).

@@

We simulated trophic community dynamics in branching river networks using a discretemixed-time model which models predation continuously and has discrete, synchronous reproduction. This is a life-history common to many lotic species, including macroinvertebrates and fish. Accounting for this stage structure is an important aspect which is often missed in continuous-time models57. We describe the model in detail below.

Our simulation model consists of three species communities; a basal species B which has positive logistic growth up to a carrying capacity, K; a primary consumer species, C which grows based on the consumption of B; and a predator P, which can consume both B and C. In our model, the realized species abundance Nix(t+1) for species i (either B, C, or P) at patch x and time t is expressed as a composite of within-patch trophic interactions, immigration, and emigration:

Nix(t+1)∼ Poisson(max{nix(t)+Iix(t)-Eix(t), 0})

Carrying capacity, KK, in patch xx , Kx ,was assumed to increase basen onexponentially as a power function of ecosystem area (i.e., watershed area)58:

Kx= Kbase+C\*N

Where KBase = 100, C is a constant, N = number of patches upstream, and ψ was a random variable sampled from Unif(1.1, 1.2). At upstream terminal nodes (headwaters) K = KBase.

where nix(t) is the expected species abudance after within-patch trophic interactions, Iix(t) the expected number of immigrants into patch x, and Eix(t) the expected number of emigrants from patch x. The realized discrete number of individuals is drawn from a Poisson distribution to account for stochasticity in demographic and dispersal processes.Let Nix(t) Nix(t) represent the population density of species i in patch x at time t. The model starts with some initial nonzero population densities for all three species in each patch. The initial population abundances for each species i in each patch x were independently sampled from a random Poisson distribution Poisson(ix)Poisson(𝝺ix), where ix𝝺ix for B, C and P was 0.8Kx \* K, 0.5Kx \*K and 0.25Kx \* K, respectively.  First, dispersal from all patches is accounted for. The number of emigrants, Eix(t) for species i in from patch x at time t, is estimated as the product of dispersal probability, Pdispersal and species abundance Nix(t). The dispersal probability was assumed to be constant for all three species in each simulation run. Then, the number of emigrants from patch x is divided into immigrants, Iix(t) Ixy(t) to each patch y. The immigration probability at patch x and time t, ix(t) xy ξxy from patch x to y is calculated considering network structure and species dispersal capability as followsbased on the habitat structure among patches according to:

>>>>

---
Title: "Ecosystem size and complexity are extrinsic drivers of food chain length in branching networks"
output:
  pdf_document:
    latex_engine: xelatex
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
bibliography: references.bib
---

@@

We generated branching networks as connected habitat patches (**Figure 1A**). The geometric arrangement of networks was controlled by two parameters: ecosystem size, the number of patches ($N_{p}$); and ecosystem complexity, branching probability ($P_{b}$). Habitat patches were assigned to be a confluence (or upstream terminal patch) with probability $P_{b}$, or an in-branch patch with probability $1-P_{b}$. Branches (tributaries) are a series of connected patches between confluences (or upstream terminal patches). The number of patches within branch $q$, $n_{p,q}$, is a random variable drawn from a geometric distribution $n_{p,q} \sim Ge(P_{b})$ but conditional on $N_{p} = \sum_{q}^{N_{b}} n_{p,q}$ ($N_{b}$ is the number of branches in a network). Full details of the network generation process is described in @teruiEmergentDualScaling2021. For each ecological scenario described below, we generated 1000 networks across a gradient of ecosystem size and complexity by drawing $N_{p}$ and $P_{b}$ from uniform distributions as: $N_{p} \sim Unif(10, 150)$ and $P_{b} \sim Unif(0.01, 0.99)$ ($N_{p}$ was rounded to the nearest integer before running simulations).

We simulated the spatial distribution of disturbance intensity (i.e., proportional mortality in the event of stochastic regional disturbance) as follows. A disturbance value, $m_{h}$, was assigned randomly to upstream terminal patch $h$ as $logit~m_{h} \sim Normal(logit~m, \sigma_{h}^2)$, where $m$ is the average of disturbance intensity at headwaters in an ordinary scale. The disturbance values propagate downstream through a spatial autoregressive process as $logit~m_{down} = logit~m_{up}+ \epsilon$, where $m_{down}$ and $m_{up}$ are the disturbance values at down- and upstream patches, and the local environmental noise at the downstream patch $\epsilon$ was drawn from $Normal(0, \sigma_{l}^2)$. This autoregressive process resembles natural disturbance regime, where flood or drought disturbance cascades downstream with some local variation along the channel. At confluence patches, the disturbance value takes a weighted mean based on the relative size of the two contributing branches, $Q$ (the number of upstream contributing patches): $logit~m_{down} = \omega~logit~m_{1,up} +(1-\omega)~logit~m_{2,up}$, where $m_{1,up}$ and $m_{2,up}$ are the disturbance values of upstream contributing tributaries, and $\omega$ the relative size of one tributary over another defined as $\omega = \frac{Q_{1}}{Q_{1} + Q_{2}}$. Thus, the larger tributary has a greater influence on the downstream environment, as observed in natural river networks [@bendaNetworkDynamicsHypothesis2004]. Disturbance value distributions had medians close to $m$ with long tails (**Figure 1B**). The low-magnitude patches can be viewed as refugia, which supply individuals to impacted sites in subsequent time steps.

We also mimicked the downstream increase of local carrying capacity [@teruiThreeEcologicalFactors2016; @carraraComplexInteractionDendritic2014; @rahelFishAssemblagesHabitat1991]. Carrying capacity in patch $x$, $K_x$, was assumed to increase as a power function of ecosystem area (i.e., watershed area) [@finlayStreamSizeHuman2011; @koenigEmergentProductivityRegimes2019]: $$
K_{x} = K_{base}N_{p,up}^z
$$where $K_{base}$ is the carrying capacity at the upstream terminal patch, $N_{p,up}$ the number of patches upstream, and $z$ is the scaling exponent. We used $K_{base} = 100$ and $z = 0.54$ based on empirical relationships of gross primary production and watershed area [@finlayStreamSizeHuman2011].

@@

We employed a discrete-time model to describe tri-trophic dynamics in each habitat patch. A discrete model is suitable for organisms with predictable reproduction cycles, a common life-history for many lotic species including macroinvertebrates and fish. Our simulation model consists of three species communities; a basal species $B$ with positive logistic growth up to a carrying capacity, $K$; a primary consumer species or intraguild prey, $C$, which grows based on the consumption of $B$; and an intraguild predator $P$, which can consume both $B$ and $C$ (**Figure 1C**). In our model, the realized species abundance $A_{ix}(t+1)$ for species i (either $B,C,~or~P$) at patch $x$ and time $t+1$ is expressed as a composite of within-patch trophic interactions, immigration, and emigration:

$$
A_{ix}(t+1) \sim Poisson(\bar{A}_{ix}(t) + I_{ix}(t) - E_{ix}(t))
$$

where $\bar{A}_{ix}$ is the expected species abudance after within-patch trophic interactions, \$I\_{ix} the expected number of immigrants into patch $x$, and $E_{ix}(t)$ the expected number of emigrants from patch $x$. The realized discrete number of individuals is drawn from a Poisson distribution to account for stochasticity in demographic and dispersal processes.

Let Nix(t) Nix(t) represent the population density of species i in patch x at time t. The model starts with some initial nonzero population densities for all three species in each patch. The initial population abundances for each species i in each patch x were independently sampled from a random Poisson distribution Poisson(ix)Poisson(𝝺ix), where ix𝝺ix for B, C and P was 0.8Kx \* K, 0.5Kx \*K and 0.25Kx \* K, respectively.  First, dispersal from all patches is accounted for. The number of emigrants, Eix(t) for species i in from patch x at time t, is estimated as the product of dispersal probability, Pdispersal and species abundance Nix(t). The dispersal probability was assumed to be constant for all three species in each simulation run. Then, the number of emigrants from patch x is divided into immigrants, Iix(t) Ixy(t) to each patch y. The immigration probability at patch x and time t, ix(t) xy ξxy from patch x to y is calculated considering network structure and species dispersal capability as followsbased on the habitat structure among patches according to:
